---
title: "Data Wrangling with AI"
author: "Matthew Mayer"
date: "2026-16-01"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
execute:
  freeze: auto
editor:
  markdown:
    wrap: 72
---

**Note**: this code book will follow the structure and content of the published code book for the project in : https://matthewmayer.quatro.pub/m04-1-principles-data-wrangling-with-tidyverse-in-r/

# overview

## Learning Outcomes

> By the end of this module you will be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions. (Next module)
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with mutate() and case_when().
-   Use the pipe operator to shape the data to prepare for analysis and visualization

## the textbook chapter to cover

-   Ch3: data Transformation
-   Ch5: Data Tidying
-   Ch13: Numbers

# introduction to Data wrangling

## Loading the packages

```{r}
library(tidyverse)
```

## What is Tidy Verse?

the tidyverse is a collection of R packages that share a common design philosophy and are designed to work together seamlessly. The tidyverse includes packages for data manipulation, visualization, and modeling, among other tasks some the core packages in the tidyverse include: - ggplot2: for data visualization - dplyr: for data manipulation - tidy r: for data tidying - readr: for data import - purrr: for functional programming - tibble: for data frames - string r: for string manipulation - formats: for working with categorical variables

## Initial Data Preperation and explantion

```{r}
library(nycflights13)
data()
```

# Data wrangling with dplyr

NOTE: use 'shift + alt +I' for multi cursor activation

- airlines Airline names
- airports Airport metadata
- flights Flights data 
- planes Plane metadata
- weather Hourly weather data

```{r}
flights %>%
  count(year, month) %>%
  arrange(desc(n)) %>%
  mutate(month = as_factor(month)) %>%
  mutate(month = fct_reorder(month, n)) %>%
  ggplot(aes(x = month, y = n, fill = factor(year))) +
  geom_col(color = "#c6891f", show.legend = FALSE) +
  coord_flip() +
  labs(
    x = "Month",
    y = "Number of flights",
    title = "# of Flights by Month during 2013 at New York"
  )
```
