---
title: "Data Wrangling with R"
author: "Allen Cosme"
date: today
format: 
    html:
        theme: cosmo
        toc: true
        toc-depth: 3
        toc-location: right-body
        embed-resources: true
        number-sections: true
execute: 
  freeze: true
---

::: callout-note
**note**: this codebook will follow the structure and content of the published codebook of 
[M05-1-Data wrangling with Tidyvverse in R](https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/) by Jae Jung, Ph.D.
:::

# Overview

## Learning outcome

> By the end of this module, you should be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions.
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with `mutate()` and `case_when()`.
-   Use the pipe operator to shape the data to prepare for analysis and visualization

## Textbook chapters covered

- Ch3: Data Transformation
- Ch5: Data Tidying
- Ch13: Numbers

# Introduction to data wrangling

## loading packages
```{r}
library(tidyverse)
```

# what is Tidyverse

The tidyverse is a collection of R packages that share a common design philosophy and are designed to work together seamlessly. The tidyverse includes packages for data manipulation, visualization, and modeling, among other tasks. 

> Some of the core packages in the tidyverse include:

-  ggplot2: for data visualization
-  dplyr: for data manipulation
-  tidyr: for data tidying
-  readr: for data import
-  purrr: for functional programming
-  tibble: for data frames
-  stringr: for string manipulation
-  forcats: for working with categorical variables

## intitial data prep and exploration

```{r}
#install.packages("nycflights13")
library(nycflights13)
data()
```

# data wrangling with dplyr

:::{.callout-note}
For MacOS, use `shift + option + I` for multi cursor activation
:::

Data sets in package ‘nycflights13’:

- airlines                Airline names.
- airports                Airport metadata
- flights                 Flights data
- planes                  Plane metadata.
- weather                 Hourly weather data

```{r}
flights <- flights

flights |>
  count(year, month) |>
  arrange(desc(n)) |>
  mutate(month = as_factor(month)) |>
  mutate(month = fct_reorder(month,n)) |>
  ggplot(aes(month, n, fill = year)) +
  geom_col(fill = "#FF007F", show.legend = FALSE) +
  coord_flip() +
  labs(x = "Month", 
      y = "Number of flights",
      title = " Number of Daily flights by month during 2013 at New york"
      )
```

